---
title: "ReproRehab Pod: Getting Started with Data in R"
subtitle: "Week 2 — Objects, Environment, Packages, and Simple Calculations"
author: "Madeline Ratoza"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Today: Let's Work With Data

Last week, we learned where things live in RStudio, how code chunks work, and why saving our code matters.

Today we will spend more time **doing R**. Our goal is to take a small dataset, bring it into R, and start answering questions about it.

## Learning goals

By the end of this session, you should be able to:

-   create and recognize objects in R;
-   use the Environment to see objects in your current session;
-   explain the difference between installing and loading a package;
-   import a `.csv` file;
-   inspect a dataset;
-   perform simple calculations using variables in a dataset.

# 1. Quick Review: Objects

An **object** is a name we give to something we want R to remember.

```{r}
my_number <- 10
my_number
```

Look at your **Environment** pane. You should now see `my_number`.

The assignment operator `<-` can be read as:

> **Save what is on the right as the name on the left.**

Try another:

```{r}
number_of_students <- 12
number_of_students
```

## Your turn

Create an object called `my_age` and assign it a number.

```{r}
# Write your code here

```

# 2. Packages: One Important Distinction

Packages add functions and tools to R.

```{r eval=FALSE}
# Do this once to install a package on your computer
install.packages("tidyverse")

# Do this when you want to use the package in a session
library(tidyverse)
```

> **Install = put it on your computer.**\
> **Library = make it available in this session.**

We will use packages much more later. For today's work, base R is enough.

# 3. Meet Our Fake Rehab Study

Today's file is `rehab_study.csv`. It is a completely **synthetic dataset** created only for learning.

Each row represents one fictional study participant. Variables include:

-   `participant_id`
-   `age`
-   `group`
-   `baseline_score`
-   `followup_score`
-   `visits`
-   `completed`

Before importing it, open the CSV outside R. **What do you notice?**

# 4. Import the Data

Make sure [`rehab_study.csv`](https://drive.google.com/file/d/1TXFdcLmC8UcCmcqVIWfOowiJB4l-797D/view?usp=drive_link) is saved in the same project/folder you are working in.

```{r}
rehab_data <- read.csv("rehab_study.csv")
```

Look at your **Environment**. What changed?

This line brings together several concepts:

-   `rehab_data` = the object we are creating
-   `<-` = assignment
-   `read.csv()` = the function
-   `"rehab_study.csv"` = the file we are asking R to read

# 5. Take a Look at the Data

We should always inspect data after importing it.

```{r}
head(rehab_data)
```

```{r}
names(rehab_data)
```

```{r}
dim(rehab_data)
```

`dim()` gives us the dimensions: **rows × columns**.

```{r}
summary(rehab_data)
```

You do not need to understand everything in the summary yet. Ask:

> Does this look like the dataset I expected to import?

# 6. Simple Calculations

We can use `$` to refer to a variable inside our dataset.

```{r}
mean(rehab_data$age)
```

Read this as: find `age` inside `rehab_data`, then calculate its mean.

Try:

```{r}
mean(rehab_data$baseline_score)
mean(rehab_data$followup_score)
min(rehab_data$visits)
max(rehab_data$visits)
median(rehab_data$age)
```

# 7. Save a Result as an Object

Instead of only printing a result, we can save it.

```{r}
average_age <- mean(rehab_data$age)
average_age
```

Look at the Environment again. You should now have both `rehab_data` and `average_age`.

> **The saved code is the recipe. The Environment is only what is currently on the counter.**

# 8. Your Turn: Mini Analysis

Work independently or with a partner.

### 1. How many participants are in the dataset?

Hint: What did `dim()` tell us?

```{r}
# Your code

```

### 2. What is the average age?

```{r}
# Your code

```

### 3. What is the average baseline score?

```{r}
# Your code

```

### 4. What is the average follow-up score?

```{r}
# Your code

```

### 5. What is the maximum number of visits?

```{r}
# Your code

```

### 6. Create an object called `mean_followup` containing the average follow-up score.

```{r}
# Your code

```

Check your Environment. Do you see `mean_followup`?

# 9. Optional Challenge

If you finish early:

```{r}
table(rehab_data$group)
```

```{r}
table(rehab_data$completed)
```

What question does each line answer?

Now try:

```{r}
mean(rehab_data$followup_score) - mean(rehab_data$baseline_score)
```

What does that number represent?

> We will learn better ways to compare and manipulate groups of data in later sessions.

# Wrap-Up

Today we went from a **CSV sitting on our computer** to a dataset we could explore and analyze in R.

Can you explain:

-   What is an object?
-   What does `<-` mean?
-   What appears in the Environment?
-   What is the difference between `install.packages()` and `library()`?
-   What does `read.csv()` do?
-   What does `$` allow us to access?
-   Why should we save the code that produced our results?

# 
